---
output:
  pdf_document: default
  html_document: default
---
# Textual
In this chapter, we will go through a number of R functions for basic statistics. We will mostly use the builtin functions (from R standard library). Extra packages will be introduced whenever necessary.

## Descriptive statistics
We are going to use builtin datasets in R. You can view the available datasets by
```{r}
data()
```

```{r}
## Data sets in package ‘datasets’:

## AirPassengers                     Monthly Airline Passenger Numbers 1949-1960
## BJsales                           Sales Data with Leading Indicator
## BJsales.lead (BJsales)            Sales Data with Leading Indicator
## BOD                               Biochemical Oxygen Demand
## CO2                               Carbon Dioxide Uptake in Grass Plants
## ...
```

View the data, for example
```{r}
women
```
View the dimension, i.e. number of subjects and variables
```{r}
dim(women)
```

Obtaining mean
```{r}
mean(women$weight)
```
and median
```{r}
median(women$weight)
```
and sd
```{r}
sd(women$weight)
```
and IQR
```{r}
IQR(women$weight)
```
There 9 types of IQR in R, the default one is type 7. You may change this to type 6 (Minitab and SPSS),
```{r}
IQR(women$weight, type = 6)
```
and minimum, maximum and range
```{r}
min(women$weight)
max(women$weight)
range(women$weight)
```

However, it is actually simpler to obtain most these in one single command for both weight and height
```{r}
summary(women)
```
even simpler, all of the statistics using *psych* package
```{r eval = FALSE}
install.packages("psych")
```

```{r}
library(psych)
describe(women)
```



## Tables

### Count, proportion, percentage and cross-tabulation
Use *birthwt* dataset from MASS package.
```{r eval=FALSE}
install.packages("MASS")
```

```{r}
library(MASS)
head(birthwt)  # First six subjects
```

Count and proportion,
```{r}
table(birthwt$smoke)
prop.table(table(birthwt$smoke))
```

Cross-tabulation of smoking vs low birth weight baby,
```{r}
table(birthwt$smoke, birthwt$low)  # without row/column labels
table("Smoking status" = birthwt$smoke, "Low birth weight" = birthwt$low)  # with row/column labels
```
To add value labels to the data for a nicer table, we use *factor*
```{r}
birthwt$smoking = factor(birthwt$smoke, levels = 0:1, labels = c("Non-smoker", "Smoker"))
birthwt$low.weight = factor(birthwt$low, levels = 0:1, labels = c("Low <2.5kg", "Normal >2.5kg"))
head(birthwt)  # we added two new variables with factors
table(birthwt$smoking)
prop.table(table(birthwt$smoking))*100  # in percent
cbind(n = table(birthwt$smoking), "%" = 100*prop.table(table(birthwt$smoking)))  # using cbind
table(birthwt$smoking, birthwt$low.weight)
```
Save table for later view and analysis,
```{r}
smoke.x.weight = table(birthwt$smoking, birthwt$low.weight)
smoke.x.weight
```

### Entering table data
```{r}
smoking = as.table(rbind(c(15, 5), c(7, 13)))
smoking
str(smoking)
dimnames(smoking) = list(
  Smoking = c("Yes", "No"),
  Lung.CA = c("Yes", "No")
)
smoking
```
