---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# Textual
In this chapter, we will go through a number of R functions for basic statistics. We will mostly use the builtin functions (from R standard library). Extra packages will be introduced whenever necessary.

## Basic descriptive statistics
We are going to use builtin datasets in R. You can view the available datasets by
```{r}
data()
```

```{r}
## Data sets in package ‘datasets’:

## AirPassengers                     Monthly Airline Passenger Numbers 1949-1960
## BJsales                           Sales Data with Leading Indicator
## BJsales.lead (BJsales)            Sales Data with Leading Indicator
## BOD                               Biochemical Oxygen Demand
## CO2                               Carbon Dioxide Uptake in Grass Plants
## ...
```

We will start by using `chickwts` data that contains both numerical (weight) and categorical (feed) variables. We can view the data,
```{r}
head(chickwts)
tail(chickwts)
```
View the dimension of the data (row and column).
```{r}
dim(chickwts)
```
Here we have 71 rows (71 subjects) and two columns (two variables).

Next, view the names of the variables,
```{r}
names(chickwts)
```
and view the details about the data,
```{r}
str(chickwts)
```
which shows that `weight` is a numerical variable and `feed` is a factor or a categorical variable. `feed` consists of six categories or levels.

We can view the levels in `feed`,
```{r}
levels(chickwts$feed)
```

### Describing a numerical variable
A numberical variable is described by a number of descriptive statistics below.

To judge the central tendency of the `weight` variable, we obtain its mean,
```{r}
mean(chickwts$weight)
```
and median,
```{r}
median(chickwts$weight)
```
To judge its spread, we can view its minimum, maximum and range values
```{r}
min(chickwts$weight)
max(chickwts$weight)
range(chickwts$weight)
```
and obtain its standard deviation (SD)
```{r}
sd(chickwts$weight)
```
variance,
```{r}
var(chickwts$weight)
```
quantile,
```{r}
quantile(chickwts$weight)
```
and interquartile range (IQR)
```{r}
IQR(chickwts$weight)
```
There are nine types of quantile algorithms in R (for `quantile` and `IQR`), the default being type 7. You may change this to type 6 (Minitab and SPSS),
```{r}
quantile(chickwts$weight, type = 6)
IQR(chickwts$weight, type = 6)
```
As an alternative to IQR, we can obtain its median absolute deviation (MAD),
```{r}
mad(chickwts$weight)
```
It is actually simpler to obtain most these in a single command,
```{r}
summary(chickwts$weight)
```
even simpler, obtain all of the statistics using `describe` in the `psych` package
```{r eval = FALSE}
install.packages("psych")
```
```{r}
library(psych)
describe(chickwts$weight)
```


### Describing a categorical variable
A categorical variable is described by its count, proportion and percentage by categories.

We obtain the count of the `feed` variable,
```{r}
summary(chickwts$feed)
table(chickwts$feed)
```
both `summary` and `table` give the same result.

`prop.table` gives the proportion of the result from the count.
```{r}
prop.table(table(chickwts$feed))
```
the result can be easily turned into percentage,
```{r}
prop.table(table(chickwts$feed))*100
```
To view the count and the percentage together, we can use `cbind`,
```{r}
cbind(n = table(chickwts$feed), "%" = prop.table(table(chickwts$feed))*100)
```
We need the quotation marks " " around the percentage sign %, because % also serves a mathematical operator in R.

### Entering table data
```{r}
smoking = as.table(rbind(c(15, 5), c(7, 13)))
smoking
str(smoking)
dimnames(smoking) = list(
  Smoking = c("Yes", "No"),
  Lung.CA = c("Yes", "No")
)
smoking
```
