# Data and variable manipulation 

# Grammar of variables 

## Prepare folder and data

### Set the working directory

This can be done in 2 ways:

1. Using codes
2. Using point and click


To use point and click, use the down arrow button next to *More* . Then click 'Set as working directory'

## Read Data

```{r}
library(foreign)
data_qol<-read.dta('qol.dta',convert.factors = T)
str(data_qol)
```

## Browse data

1.  First few rows
2.  Last few rows

```{r}
head(data_qol)
tail(data_qol)
```

## Grammar of variables 

### Select columns

## Select columns

Let us create a new dataframe with only id, sex and hba1c as the variables

```{r}
data_qol2<-subset(data_qol, select = c('sex', 'age', 'hba1c'))
str(data_qol2)
```

alternatively, we can use other subsetting functions

```{r}
data_qol3<-data_qol[,c('sex','age','hba1c')]
str(data_qol3)
```

### Select rows

```{r}
data_qol4<-subset(data_qol, age > 30)
str(data_qol4)
summary(data_qol4$age)
```

alternatively, we can use other subsetting functions

```{r}
data_qol5<-data_qol[data_qol$age>30,]
str(data_qol5)
summary(data_qol5$age)
```

### Select rows and columns together

```{r}
data_qol6<-subset(data_qol,age>30 & sex=='male', select = c(id, sex, age, group))
str(data_qol6)
table(data_qol6$sex)
```

### Generate a new variable

```{r}
data_qol$age_cat<-data_qol$age
View(data_qol)
```

### Categorize into new variables

#### From a numerical variable

```{r}
data_qol$age_cat<-cut(data_qol$age_cat,
                      breaks=c(min(data_qol$age),40,60,Inf),
                      labels=c('min-39','40-59','60-above'))
min(data_qol$age)
table(data_qol$age_cat)
str(data_qol$age_cat)
```

#### From a categorical variable

```{r}
table(data_qol$tx)
str(data_qol$tx)
```

Create a variable with 'Diet only' vs 'Diet+Drug'. This is a little bit complicated

```{r}
data_qol$tx2<-data_qol$tx
str(data_qol$tx2)
str(data_qol$tx)
table(data_qol$tx2)

library(plyr)
data_qol$tx2<-revalue(data_qol$tx,c('diet only'='diet', 'OHA and diet only'='med',
                                    'insulin and diet only'='med', 'all'='med'))
table(data_qol$tx2)
```


### Dealing with missing data

```{r}
data_qol$tx3<-data_qol$tx
str(data_qol$tx3)
str(data_qol$tx)
table(data_qol$tx3)
```

#### Replace values with 'NA'

```{r}
data_qol$tx3<-revalue(data_qol$tx,c('diet only'=NA))
table(data_qol$tx3)
str(data_qol$tx3)
```

<<<<<<< HEAD
## Additional packages
=======
## Additional package
>>>>>>> b1e7fe217da24c9c49c585f8cb7b29c3a03e88ce

### Package 'dplyr'

'dplyr' package is a very useful package that encourage users to use proper verb when manipulating variables (columns) and observations (rows)

It has 9 useful functions
1.  filter()
2.  arrange()
3.  select()
4.  distinct()
5.  mutate() and transmute()
6.  summarise()
7.  sample_n() and sample_frac()

Package 'dplyr' is very useful when it is combined with another function that is 'group_by'

`



