# Reading statistical data

## Objectives

At the end of the lecture, participants are:

1.  able to read various data formats into R
2.	able to export data from R into various data format
3.	able to create R-markdown document 


## Data formats 

R can read datasets from different formats.

The `base` package enables us to read `.txt` and `.csv` files. 

You can also use the point and click method to read data. Go to File -> Import Dataset -> From ...

To read datasets from SPSS (`.sav`), Stata (`.dta`), Excel (`xlsx`) and SAS, we need to load special packages. 

There are more than 1 packages you can use to read and write data from/to different spreadsheet or statistical software.

The packages include:

1.  `haven` 
2.  `foreign`
3.  `readxl`
4.  `readr`

## Reading data into R

### Reading csv file

Let us read a `.csv` files names as `metab.csv`.

We would like to create an object named as `met_data` to represent the read `metab.csv`. We can do this: 

```{r}
met_data <- read.csv('metab1.csv', header = TRUE)
```

header = TRUE means that you will read the first row as the variable names. 

After you have done this, you will see a new object named as `met_data` inside the *environment* pane. 


### Reading dataset from MS Excel file 

You can read dataset from Excel using 2 methods

1. point and click method. File - Import Dataset - from Excel
2. `readxl` package

To read data using specific packages such as `readxl` package:

1.  you need to install the library first using `install.packages()`
2.  After that, load the library using `library(readxl)`
3.  Type `read_excel()` with relevant arguments to read the data into RStudio

```{r , eval=TRUE, echo=FALSE}
#install.packages('readxl')

```


```{r}
library(readxl)
dataexcel <- read_excel('eye.xlsx', sheet = 1)
```

The example above means that we read an MS Excel file named as  `eye.xlsx` and named it as `dataexcel`.


### Reading dataset from SPSS file (.sav) 

Dataset in SPSS format will end with `.sav`. 

To read SPSS data into R we may use `foreign` or `haven` packages. 

After reading the `.sav` file, an object will be created based on what we named it. 

The example below shows that an object named as `dataSPSS`represents the SPSS data `cholest.sav`that we just read into RStudio.  

```{r}
library(foreign)
dataSPSS <- read.spss('cholest.sav', to.data.frame = TRUE)
```

### Reading dataset from Stata (.dta)

Dataset in Stata format will end with `.dta`. 

To read Stata data into R we may use `foreign` or `haven` packages. 

After reading the `.dta` file, similarly an object will be created - in this case, we named it as `dataSTATA`. 

`dataSTATA` is an object that represents the Stata data (now in the memory) that we just read `metab1.dta`into RStudio.


```{r}
datastata <- read.dta('metab1.dta', convert.factors = TRUE)
```

### Alternative methods or package

1.  You can go to File - Read Datasets
2.  You may use **haven** package to read SAS, SPSS and Stata file.

```{r}
library(haven)
dataSPSS2 <- read_sav('cholest.sav')
dataSTATA2 <- read_dta('cholest.dta')
```

### Other data format

Other important data formats that might be useful in epidemiology and statistics are:

1.  shapefile `.shp` 
2.  text file `.txt`
3.  text file `.dat`
4.  `XML` file
5.  images file `DICOM`

We will not cover these today. 

## Exporting data from R

You can also export data into various formats using similar packages.

For example, 

1.  to export data into a *comma separated version* (.csv) file, we can use `write.csv` function. 
2.  to export data into stata format, we can use `write.dta` function

```{r}
export_csv <- write.csv(dataSPSS, 'export_csv.csv')
export_stata <- write.dta(dataexcel, 'export_stata.dta') 
```

