# GLM


### Objectives

At the end of the lecture, participants are

1.  able to perform linear regression
2.  able to perform logistic regression
3.  able to perform Cox proportional hazard regression

## Set working directory

Set your working directory. 

This is a folder that contains your dataset and objects created by R.


## Read data

We will read a stata data into R. This file will read using **foreign** package.

We will name the object as **data1** as an object that represent the dataset.

This object is a `data.frame` object

The **data1** object will remain in the memory unless you close your RStudio. 

```{r}
library(foreign)
data1 <- read.dta('metab1.dta', convert.factors = TRUE) 
head(data1)
```

We use `head()` function to list the first 6 observations in the dataset.


## Explore and clean data

Next we will describe the data and visualize 

1.  Descriptive
2.  Visualization

### Descriptive analysis

We will check basic descriptive statistics from our data

```{r}
library(psych)
describe(data1)
```


### Visualization

We do not have time to cover for data vizualition. 

But for here we would do

1.  histogram
2.  bar charts
3.  box-plots
4.  scatterplots

To examine the distribution of our data. 

Briefly:

```{r}
hist(data1$age)
```

```{r}
cts_sex <- table(data1$sex)
barplot(cts_sex, names.arg = c('male','female'))
```

```{r}
cor(data1[,c(2,7:14 )], use = 'complete.obs')
```


## Linear regression

We perform linear regression when we assume that distribution of the outcome variables is normally distributed as a function of certain covariates (independent variables)

### Estimation

To perform the estimation for linear regression, we can use `lm()` function. 

Let us model body mass index *bmi* as a function of hip circumference, mean systolic blood pressure *msbp*, mean diastolic blood pressure *mdbp* and *gender*

```{r}
modl <- lm(bmi2 ~ hip + msbp + mdbp 
           + gender, data = data1)
summary(modl)
```

From the results, we can see that:

1.  71.5% of variation in the expected *bmi* is explained by the covariates
2.  all covariates are siginificantly (p < 0.05) predictive of *bmi*

### Inference

Now, let us calculate the 95% of the expected mean of bmi

```{r}
confint(modl)
```

### Prediction


## Logistic regression

In logistic regression, we model an outcome variables which is assumed to follow binomial distribution as a function of a set of covariates (independent variables).

In R, we use `glm()` function to perform **Generalized Linear Regression** analysis. 

But there are other R packages that can do similar analysis. Based on our experience, the `glm()` in the **base** package is good enough.

### Estimation

Let us estimate the expected log odds for having diabetes mellitus *dm* as a function os these covariates: body mass index *bmi2* , age , total cholesterol *totchol *and mean systolic blood pressure *msbp*. 

We specify the family of *binomial* and use the *logit* link.

```{r}
modlog <- glm(dm ~ bmi2 + age + totchol + msbp, family = binomial(link = logit), data = data1)
summary(modlog)
```

From the results, we estimate that the log odds for diabetes mellitus:

1.  increase by 0.049 for each increase in bmi, adjusted for other covariates (p-value 0.072)
2.  increase by 0.048 for each increase in age, controlled for other covariates (p-value < 0.001) 
3.  are not significantly predicted by either the total cholesterol and mean systolic blood pressure (p-value 0.717 and 0.572, respectively) 

Odds ratio

To obtain the odds ratios, we use the `exp()` function

We then exponential the coefficients of the model estimated from `glm()` function

```{r}
exp(modlog$coefficients)
```

### Inference

To estimate the 95% confidence intervals for the log odds (CI for log odds), we use the `confint()` function to all the regression parameters or the $\beta_p$

```{r}
confint(modlog)
```

And to estimate the CI for odds ratios, we exponentiate the lower bound and upper bound of the regression parameters $\beta_p$

```{r}
exp(confint(modlog))
```


## Cox proportional hazard regression

In a cohort study, a time-to-event data is common. Such data has a time from the start of study until one point of time. 

This point of time can be either:

1.  participant develop an outcome of interest
2.  participant do not develop the outcome of interest after the maximum time of follow up

In R, the most common packacge for typical time-to-event data is **survival** package

To perform time-to-event data analysis, we need to load the **survival** library.

We will use a data frame using the built-in lung cancer dataset that ships with the survival package. :

1.  inst: Institution code
2.  time: Survival time in days
3.  status: censoring status 1=censored, 2=dead
4.  age: Age in years
5.  sex: Male=1 Female=2
6.  ph.ecog: ECOG performance score (0=good 5=dead)
7.  ph.karno: Karnofsky performance score as rated by physician
8.  pat.karno: Karnofsky performance score as rated by patient
9.  meal.cal: Calories consumed at meals
10. wt.loss: Weight loss in last six months


Load the library  

```{r}
library(survival)

```

Describe data

```{r}
library(psych)
lung <- lung
describe(lung)
```

and declare the time to event

```{r}
tte <- Surv(time = lung$time, event = lung$status==2)
tte
```

### Estimation and inference

perform Cox PH model to estime the log hazard and the hazard ratios.

1.  constant only model

```{r}
cox_mod <- coxph(tte ~ 1, data = lung)
summary(cox_mod)
```

2.  multivariable model

```{r}
cox_mod2 <- coxph(tte ~ 1 + ph.ecog + factor(sex) + ph.karno, data = lung)
summary(cox_mod2)

```



